---
title: "Assignment 12 - Connectivity"
output: html_notebook
---


```{R}
rm(list=ls())

require(tidyverse)
require(terra)
require(sf)
require(gdistance)
require(igraph)
require(fitdistrplus)
require(fdrtool)
require(raster)
require(Hmisc)



#read in data

# Read in the landcover raster
land = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week10/panther_landcover.tif')

# Identify the crs of the landcover raster
crs(land)

#Look at th raster's resolution
res(land)



# Bring in the public lands shapefile
public = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week12/panther_publicland.shp')

# Set the crs of the public lands shapefile
st_crs(public) = crs(land)

# And look at it
data.frame(public)


#calc centroids
publicCentroids = st_centroid(public)

#plot(land)
#plot(st_geometry(public), add=T)
#plot(st_geometry(publicCentroids), add=T, col='red')


#resistance
classification = read.table('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/resistance%20reclass.txt', header=T)
landCost = classify(land, classification[,c(1,3)])
landCost = raster(landCost)

#plot(landCost)
#plot(st_geometry(public), add=T)
#plot(st_geometry(publicCentroids), add=T, col='red')
```


# Challenge 1 (5 points)

In the lab, we calculated Euclidean distance, least-cost distance, commute distance (based on circuit theory), and a randomized shortest path distance. Examine the correlation among these 4 metrics of effective distance. Which tend to be the longest, which tend to be the shortest, and why? In what situations might you expect to find strong differences between Euclidean distance and the other metrics of effective distance?

```{r}
#Code here

#Euclidean distance

geoDist = pointDistance(publicCentroids, lonlat=F)
geoDist = as.dist(geoDist)
geoDistvect = as.vector(geoDist)

#Least-Cost distance

# Calculate conductance transition matrix
landCond = transition(1/landCost, transitionFunction = mean, 8)

#Do a geographic correction of the conductance values to "correct" for the fact that there are further distances between cells that touch on the diagonal
landCond = geoCorrection(landCond, type='c', multpl=F)
lcDist = costDistance(landCond, st_coordinates(publicCentroids))
lcDist
lcDistVect = as.vector(lcDist)

#Commute distance
circuitDist = commuteDistance(landCond, st_coordinates(publicCentroids))
circuitDist
circuitDistVect = as.vector(circuitDist)

#Randomized shortest path distance
rspDist1.2 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[2,], theta=0.001)
rspDist1.2

rspDist1.3 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[3,], theta=0.001)

rspDist1.4 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[4,], theta=0.001)

rspDist1.5 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[5,], theta=0.001)

rspDist2.3 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[2,], to=st_coordinates(publicCentroids)[3,], theta=0.001)

rspDist2.4 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[2,], to=st_coordinates(publicCentroids)[4,], theta=0.001)

rspDist2.5 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[2,], to=st_coordinates(publicCentroids)[5,], theta=0.001)

rspDist3.4 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[3,], to=st_coordinates(publicCentroids)[4,], theta=0.001)

rspDist3.5 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[3,], to=st_coordinates(publicCentroids)[5,], theta=0.001)

#not working
rspDist4.5 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[4,], to=st_coordinates(publicCentroids)[5,], theta=0.001)

rspDistVect = c(rspDist1.2, rspDist1.3, rspDist1.4, rspDist1.5, rspDist2.3, '0', rspDist2.5, rspDist3.4, rspDist3.5, '0')

```


Now to figure out the correlations

```{R}
#Correlation

#create a vector of each matrix, add that as a column in a new dataframe, and run a correlation.chart between all of the columns in the dataframe. 

comparisons = c('1.2', '1.3', '1.4', '1.5', '2.3', '2.4', '2.5', '3.4', '3.5', '4.5')
#dataframe with 
Distances2 = cbind(comparisons, geoDistvect, lcDistVect, circuitDistVect, rspDistVect)
#Distances2 = as.data.frame(Distances2)

#Distances2 = as.numeric(Distances2)

Distances2
rcorr(as.matrix(Distances2))

#Couldn't get either of these two to work, so just used a different method for getting correlations
#PerformanceAnalytics::chart.Correlation(Distances2, histogram=F)

#Distances2 %>% 
#  as.data.frame() %>% 
#  select(comparisons, geoDistvect, lcDistVect, circuitDistVect, rspDistVect) %>% 
#  PerformanceAnalytics::chart.Correlation(histogram=F)


```

Circuit (commute distance) is the longest and the geoDist tends to be the shortest. It makes sense that the geoDist is the shortest as it is going to calculate the shortest Euclidian distance without taking into effect resistance across the landscape, so it will always find the shortest path even though that might not be realistic as the lanscape surface may vary and introduce resistance. The circuit (commute) being the longest makes sense because it is a random walk, which is going to innately introduce much longer paths than following the shortest distance. 

I would expect strong differences between Euclidean distance and metric of effective distance when the landscape has areas of high resistance, whereas when there is little to no resistance on the landscape, I anticipate that the effective distances will begin to approach the Euclidean shortest distance, as the resistance matrix would be effectively zero and not increase the distance values. The instances where there would still be a difference are likely in the circuit and random walk methods as they both include random walk components that will naturally deviate from the shortest euclidean path. 


# Challenge 2 (5 points)

In the lab we developed a 10% least-cost corridor between two protected areas. Identify what proportion of this corridor is comprised of each landcover type based on the land raster. To do this, make sure that both the land and leastCostCorridor10 rasters are spatRasters. Then mask land with leastCostCorridor10. The values in the resulting raster will be representative of the landcover values in the corridor. Summarise this information and refer back to the classification dataframe to link the landcover numbers with landcover types. What are the three most common landcover types within this corridor, and does this make sense given what you know about the resistance to movement within such landcover types?

```{r}
#Code here

#mapping least-cost paths
# Choose the extent for cropping
fpwrOssfExtent = extent(642000, 683000, 237000, 298000)

# Crop the landcover layer
landSub = crop(land, fpwrOssfExtent)

# Crop the cost layer and create a transition matrix
landCostSub = crop(landCost, fpwrOssfExtent)
landCondSub = transition(1/landCostSub, transitionFunction=mean, 8)
landCondSub = geoCorrection(landCondSub, type='c', multpl=F)



#actually map
fpwrOssfLcp = shortestPath(landCond, st_coordinates(publicCentroids)[5,], st_coordinates(publicCentroids[3,]), output='SpatialLines')

plot(landCostSub, axes=F)
plot(st_geometry(public), add=T)
plot(st_geometry(publicCentroids), col='grey30', add=T)
lines(fpwrOssfLcp, col='red', lwd=3)

#CORRIDOR
fpwrCost = accCost(landCondSub, st_coordinates(publicCentroids)[5,])
ossfCost = accCost(landCondSub, st_coordinates(publicCentroids)[3,])

plot(fpwrCost)
plot(ossfCost)

leastCostCorridor = overlay(fpwrCost, ossfCost, fun=function(x, y){return(x+y)})

plot(leastCostCorridor)

#Calculate the 10% quantile for the leastCostCorridor raster
quantile10 = quantile(leastCostCorridor, probs=0.1, na.rm=T)
quantile10

#Create a new raster with a value of 1 if it has a summed cost in the lower 10% and a NA otherwise
leastCostCorridor10 = leastCostCorridor
values(leastCostCorridor10) = NA
leastCostCorridor10[leastCostCorridor < quantile10] = 1

#Plot this raster and look at it
plot(leastCostCorridor10, legend=F, axes=F)
points(publicCentroids, col='grey30')
lines(fpwrOssfLcp, col='red')


#help from jonathon
#make sure spatial raster
leastCostCorridor10.2 = rast(leastCostCorridor10)
crs(leastCostCorridor10.2) = crs(land)
#land = rast(land)

corridorComp = crop(land, leastCostCorridor10.2)
corridorComp = mask(corridorComp, leastCostCorridor10.2)
table(matrix(corridorComp))

prop.table(table(as.vector(corridorComp)))[1:20]


#now take this and look at our classification dataframe to match landcover type with the corresponding resistance value
classify

#These lines are to reclassify the landcover type into a resistance value if necessary - but I didn't need it to answer the question
#corridorCompRes = classify(corridorComp, classification[,c(1,3)])
#corridorCompRes = raster(corridorCompClass)
#plot(corridorCompRes)

```


The three most common landcover types are 17-cypress swamp at a proportion of .22 which has a resistance value of 3, 9- pinelands at a proportion of 0.143 which has a resistance value of 2, and 12- Freshwater marsh at a proportion of .137 which has a resistance value of 5. 

This makes sense that the more common landcover within the 10% least-cost corridor have low resistance values, as the corridor is calculated based off of a least-cost path which is determined based off of the provided resistance values. It is interesting that the most common landcover type has a higher resistance value than the second most common, however that is likely just by chance of what exists on the landscape. 


# Challenge 3 (5 points)

In the lab, we used the passage() function to conduct a randomized shortest path analysis mapping connectivity among two protected areas. We initially set theta to 0 thus resulting in a circuit theory based map of connectivity. Using the same passage() function, conduct the analysis again using values of theta = 0.0001 and theta = 0.001. Plot the resulting maps. What patterns do you see emerge as theta increases? Comment on the tradeoffs between biological realism and ease of conservation planning based on these three different maps of connectivity.

```{r}
#Code here from class
#Randomized shortest path

#Random walk

passageMapT0 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0)

passageMapT0.0001 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.0001)

passageMapT0.001 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.001)

plot(passageMapT0)
plot(passageMapT0.0001)
plot(passageMapT0.001)

```

As theta increases, the random walk becomes less random and more of the random walks follow the same path (which resembles the least-costs path). The more sporadic paths are likely more realistic for an organism, as it isn't necessarily going to follow the least-cost path, and there is some randomness in every individual, along with different complex choices that lead to a path, that would cause them to stray from the least-cost path. However, the more random path (lowest theta) is also more spread out and covers essentially the entire map between the two points, which is not realistic for conservation planning. Conservation planning is going to be much easier with a narrower area (the highest theta) which resembles the least-costs path and is the least realistic for planning actual movements of individual organisms. Going with an in between value may be the best bet to make it easier for conservation planning, yet more realistic in terms of the randomness and high likelihood that organisms will wander off of the least costs corridor that the highest theta mimics. 


# Challenge 4 (5 points)

In the latter part of the lab, we discussed calculating patch-based and landscape-based metrics of connectivity. Patch number 6 had the lowest degree (n = 2) while patch number 7 had one of the highest degrees (n = 12). First delete patch number 6 from the network, and then examine the impacts on the landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC). Now replace patch 6 back into the network and instead delete patch number 7. Again, examine how ICC and PC change. How does the loss of patch 6 compare to the loss of patch 7 in terms of its impacts on landscape-level connectivity?

```{r}
#Code here from class

#working with a completely different dataset and comparing potential vs. realized connectivity among 29 sites for the endangered Snail Kite. First we're going to bring in a node-specific data frame that has name, coordinate, and area information for the 29 sites.
nodes = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_nodes.csv')
area = nodes$area
#nodes

#Next we will bring in a matrix showing within-breeding-season dispersal. The values in the matrix represent the observed number of Snail Kites that moved between each combination of sites. Note that this is not a symmetric matrix, meaning that we have bi-directional information.
#Note that we already have one directed, weighted network in aObs
aObs = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_movement.csv')[,-1]
diag(aObs) = 0
#aObs

coords = cbind(nodes$XCoord, nodes$YCoord)
distMat = pointDistance(coords, lonlat=F)
distMat = distMat / 1000 # in km

linkLoc = which(aObs > 0, arr.ind=T)
withinDisp = cbind(distMat[linkLoc], aObs[linkLoc])

#Creating one distance observation for each individual
withinDisp = rep(withinDisp[,1], withinDisp[,2])

#calc mean dispersal distance
aMeanDist = mean(withinDisp)

#transition matrix
aMean = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
aMean[distMat < aMeanDist] = 1
diag(aMean) = 0
#
aProb = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
alpha = 1/aMeanDist
aProb = exp(-alpha*distMat)
diag(aProb)=0
graphAmean = graph_from_adjacency_matrix(aMean, mode='undirected')
graphAprob = graph_from_adjacency_matrix(aProb, mode='undirected', weighted=T)
graphAobs = graph_from_adjacency_matrix(as.matrix(aObs), mode='directed', weighted=T)

#Patch-scale connectivity
aMeanDegree = degree(graphAmean)
aMeanStrength = strength(graphAmean)
aMeanEigen = eigen_centrality(graphAmean)
aMeanClose = closeness(graphAmean)
aProbBetween = betweenness(graphAprob, weights=1/E(graphAprob)$weight)

#landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC)
connectance = edge_density(graphAmean)

#IIC
AL = 63990 # Approximate study area in km^2
#Create a matrix of shortet paths
nlMat = distances(graphAmean)
#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat [is.infinite(nlMat)] = 1000
iicMat = outer(area, area)/(1+nlMat)
iic = sum(iicMat)/AL^2
iic


#PC
pStarMat = distances(graphAprob, weights=E(graphAprob)$weight)
# Back-transform to probabilities
pStarMat = exp(-pStarMat)
# Numerator of PC
pcNum = outer(area, area)*pStarMat
# Probability of connectivity
pc = sum(pcNum)/AL^2
pc
```


```{R}
#Code for assignment
########################
#Removing patch number 6
########################
nodes_subset = nodes[-6,] #removing patch number 6
area = nodes_subset$area

aObs_subset = aObs[-6,-6]

coords = cbind(nodes_subset$XCoord, nodes_subset$YCoord)
distMat = pointDistance(coords, lonlat=F)
distMat = distMat / 1000 # in km

linkLoc = which(aObs_subset > 0, arr.ind=T)
withinDisp = cbind(distMat[linkLoc], aObs_subset[linkLoc])

#Creating one distance observation for each individual
withinDisp = rep(withinDisp[,1], withinDisp[,2])

#calc mean dispersal distance
aMeanDist = mean(withinDisp)

#transition matrix
aMean = matrix(0, nrow=nrow(aObs_subset), ncol=ncol(aObs_subset))
aMean[distMat < aMeanDist] = 1
diag(aMean) = 0
#
aProb = matrix(0, nrow=nrow(aObs_subset), ncol=ncol(aObs_subset))
alpha = 1/aMeanDist
aProb = exp(-alpha*distMat)
diag(aProb)=0
graphAmean = graph_from_adjacency_matrix(aMean, mode='undirected')
graphAprob = graph_from_adjacency_matrix(aProb, mode='undirected', weighted=T)
graphAobs = graph_from_adjacency_matrix(as.matrix(aObs_subset), mode='directed', weighted=T)

#Patch-scale connectivity
aMeanDegree = degree(graphAmean)
aMeanStrength = strength(graphAmean)
aMeanEigen = eigen_centrality(graphAmean)
aMeanClose = closeness(graphAmean)
aProbBetween = betweenness(graphAprob, weights=1/E(graphAprob)$weight)


#landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC)
connectance = edge_density(graphAmean)

#IIC
AL = 63990 # Approximate study area in km^2
#Create a matrix of shortet paths
nlMat = distances(graphAmean)
#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat [is.infinite(nlMat)] = 1000
iicMat = outer(area, area)/(1+nlMat)
iic = sum(iicMat)/AL^2
iic

#PC
pStarMat = distances(graphAprob, weights=E(graphAprob)$weight)
# Back-transform to probabilities
pStarMat = exp(-pStarMat)
# Numerator of PC
pcNum = outer(area, area)*pStarMat
# Probability of connectivity
pc = sum(pcNum)/AL^2
pc
```


```{R}

########################
#Removing patch number 7
########################
nodes_subset = nodes[-7,] #removing patch number 6
area = nodes_subset$area

aObs_subset = aObs[-7,-7]

coords = cbind(nodes_subset$XCoord, nodes_subset$YCoord)
distMat = pointDistance(coords, lonlat=F)
distMat = distMat / 1000 # in km

linkLoc = which(aObs_subset > 0, arr.ind=T)
withinDisp = cbind(distMat[linkLoc], aObs_subset[linkLoc])

#Creating one distance observation for each individual
withinDisp = rep(withinDisp[,1], withinDisp[,2])

#calc mean dispersal distance
aMeanDist = mean(withinDisp)

#transition matrix
aMean = matrix(0, nrow=nrow(aObs_subset), ncol=ncol(aObs_subset))
aMean[distMat < aMeanDist] = 1
diag(aMean) = 0
#
aProb = matrix(0, nrow=nrow(aObs_subset), ncol=ncol(aObs_subset))
alpha = 1/aMeanDist
aProb = exp(-alpha*distMat)
diag(aProb)=0
graphAmean = graph_from_adjacency_matrix(aMean, mode='undirected')
graphAprob = graph_from_adjacency_matrix(aProb, mode='undirected', weighted=T)
graphAobs = graph_from_adjacency_matrix(as.matrix(aObs_subset), mode='directed', weighted=T)

#Patch-scale connectivity
aMeanDegree = degree(graphAmean)
aMeanStrength = strength(graphAmean)
aMeanEigen = eigen_centrality(graphAmean)
aMeanClose = closeness(graphAmean)
aProbBetween = betweenness(graphAprob, weights=1/E(graphAprob)$weight)


#landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC)
connectance = edge_density(graphAmean)

#IIC
AL = 63990 # Approximate study area in km^2
#Create a matrix of shortet paths
nlMat = distances(graphAmean)
#In cases where you have isolated patches, we assign an arbitrarily large value
nlMat [is.infinite(nlMat)] = 1000
iicMat = outer(area, area)/(1+nlMat)
iic = sum(iicMat)/AL^2
iic

#PC
pStarMat = distances(graphAprob, weights=E(graphAprob)$weight)
# Back-transform to probabilities
pStarMat = exp(-pStarMat)
# Numerator of PC
pcNum = outer(area, area)*pStarMat
# Probability of connectivity
pc = sum(pcNum)/AL^2
pc
```

The IIC and PC is lowest when removing patch 7, but not much different at all from the original dataset or when removing patch 6. 
IIC original = 0.002805411, removing 6 = 0.002799433, removing 7 = 0.002650867
PC original = 0.00595698, removing 6 = 0.005961139, removing 7 = 0.005632307
It makes sense that removing patch 7 had a bigger effect and decreased the IIC and PC more than patch 6 or the original, as patch 7 had the highest degree of connectivity amongst all of the patches. 

However, all of these values are essentially 0, indicating low connectivity and not much habitat. 
