---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
# Week 2 Lab Assignment - Claire Tracy

## Overview
I study the sergeant major damselfish (Abudefduf saxatilis) and the panamanian damselfish (Abudefduf troschelii). For this assignment I am mapping their entire ranges with sea surface temperature of the ocean to show the extents of their range and how that may relate with the environment. This rmd file will produce a zoomed in and zoomed out map showing the distribution of the two species obtained from the IUCN redlist (which has nice shapefiles for species distributions), occurrence data of both species (black points) downloaded from iNaturalist (to show how their ranges overly with publicly available occurrence data), and ocean sea surface temperature (SST) obtained from NOAA. The raster layer is the SST, and my vector elements include the shape files of the range distributions of both species (each an independent shape file), and the occurrence data of both species. 


A third layer that I am interested in adding in the future is a boundary/depiction of major ocean currents (i.e. gulf stream). Data is available in the github and was downloaded from here: https://hub.arcgis.com/maps/beyondmaps::major-ocean-currents/about. I however did not have more time to do it for this assignment so maybe in the future. 
Major_Ocean_Currents_arrowPolys_30m_8


## Getting set up
```{r}
#I am assuming that this needs all of the packages required in the Rmd file from lab. 
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)
require(rworldmap)
require(lattice)
require(raster)
require(ggOceanMaps)
require(rasterVis)
```



## Read in all the data
YOU WILL NEED TO CHANGE THE PATH ON LINES 52 AND 58, THOSE DATASETS DID NOT WORK TO PULL FROM GITHUB, BUT I LEFT THE ORIGINAL CODE BELOW COMMENTED OUT. 
```{R}

# Read in species distribution data downloaded from IUCN redlist
A_sax_dist = st_read('/vsicurl/https://raw.githubusercontent.com/C-Tracy/SpatialEcology/main/Week2/redlist_species_data_7110df13-c1f4-4b2f-9342-598f925c5d03/data_0.shp')

A_tro_dist = st_read('/vsicurl/https://raw.githubusercontent.com/C-Tracy/SpatialEcology/main/Week2/redlist_species_data_48c00095-da02-43fa-a90d-5f9a41331177/data_0.shp')

#read in inaturalist locality data - filtered to only research grade ids
#occ = read.csv('~/Downloads/Abudefduf_sax_tro_vertnet-e7736bf57d094efdbc063509f61b2425.tsv', sep='\t')


#CHANGE TO YOUR DIRECTORY IF NOT RUNNING FROM DOWNLOADS FOLDER
occ_sax = read.csv('~/Downloads/observations-396470.csv')
#Below is the command that should work pulling from github but isn't (the file isn't too large)
#occ_sax = read.csv('/vsicurl/https://raw.githubusercontent.com/C-Tracy/SpatialEcology/main/Week2/observations-396470.csv')


#CHANGE TO YOUR DIRECTORY IF NOT RUNNING FROM DOWNLOADS FOLDER
occ_tro = read.csv('~/Downloads/observations-396472.csv')
#Below is the command that should work pulling from github but isn't (the file isn't too large)
#occ_tro = read.csv('/vsicurl/https://raw.githubusercontent.com/C-Tracy/SpatialEcology/main/Week2/observations-396472.csv')

occ_comb = rbind(occ_sax, occ_tro)

occ_comb_sf = st_as_sf(occ_comb, coords=c('longitude', 'latitude'), na.fail = FALSE, crs='+proj=longlat +datum=WGS84 +units=m +no_defs +type=crs')

#occ_sax_Sf = st_as_sf(occ_sax, coords=c('longitude', 'latitude'), na.fail = FALSE, crs='+proj=longlat +datum=WGS84 +units=m +no_defs +type=crs')

#occ_tro_Sf = st_as_sf(occ_tro, coords=c('longitude', 'latitude'), na.fail = FALSE, crs='+proj=longlat +datum=WGS84 +units=m +no_defs +type=crs')


#plot(st_geometry(occ_comb_sf))

#create bbox based on locality records from inat
studyArea = st_bbox(occ_comb_sf)
studyArea = studyArea +  c(-10, -10, 10, 10)
studyArea = st_as_sfc(studyArea)

#checking projections
crs(occ_comb_sf, proj=T)
crs(A_tro_dist, proj=T)



#this command produces a nice map including bathymetry data of ocean if you don't need to add in a raster with SST, so I'm not taking it out of the script because I may use it in the future, but commented it out because it's not needed for this
#basemap(limits = c(-15, -135, -50, 50), bathymetry = TRUE) + geom_sf(data = A_sax_dist, aes(color = "red")) + geom_sf(data = A_tro_dist, aes(color = "blue"))



#Adding box around each study area
#fill can be changed to the indicated color for "color"

#basemap(limits = c(-135, 45, -50, 50), bathymetry = TRUE) + 
#  geom_sf(data = A_tro_dist, fill = '#795695', color = '#795695') + 
#  geom_sf(data = A_sax_dist, fill = '#eb6262', color = '#eb6262') +
#  geom_sf(data = occ_comb_sf, fill = "black", color = "black")


```






## Adding raster layer to incorporate sea surface temperature (SST), and plotting zoomed out map with extent rectangle.
```{r}

#basemap(limits = c(-135, 45, -50, 50), bathymetry = FALSE) + 
#  geom_sf(data = studyArea, fill=NA, color = "black") +
#  geom_sf(data = A_tro_dist, fill = NA, color = '#795695') + 
#  geom_sf(data = A_sax_dist, fill = NA, color = '#eb6262')
  


#trying a new sst raster from https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.nodc:0077816
sst_2 <- rast('/vsicurl/https://raw.githubusercontent.com/C-Tracy/SpatialEcology/main/Week2/2009-04_Pathv5.0_daynight_SST_monaprc.tif')
#plot(sst_2)

sst2_crop <- crop(sst_2, studyArea)
#plot(sst2_crop)

#use ggplot to create plot zoomed out to whole world (using non-cropped objects) and including a bbox for the study area to indicate the area of the zoomed in plot
ggplot() +
  geom_spatraster(data = sst_2) +
  scale_fill_gradient2(low='lightblue', mid = 'orange', high='darkred', midpoint = 15) +
  geom_sf(data = A_sax_dist, fill = '#eb6262', color = '#eb6262') + 
  geom_sf(data = A_tro_dist, fill = '#795695', color = '#795695') +
  geom_sf(data = occ_comb_sf, fill = "black", color = "black", size = 0.01) +
  geom_sf(data = studyArea, fill = NA, color = "black") + 
  guides(fill = guide_legend(title = "Sea Surface Temperature (C)")) 

```




## Zoomed in map

This has an interpretable legend, but I spent an hour trying to incorporate a second legend for the two distributions and the point data, but with no luck. I've decided the point is not worth the hassle and I would love your feedback if you have a tip on how to add another legend that shows the colors of the other shape files I incorporated. 
```{r}
#use ggplot to create plot zoomed into study area
ggplot() + 
  geom_spatraster(data = sst2_crop) +
  scale_fill_gradient2(low='lightblue', mid='orange', high='darkred', midpoint = 15) +
  geom_sf(data = A_sax_dist, aes(color = SCI_NAME), fill = '#eb6262', color = '#eb6262') + 
  geom_sf(data = A_tro_dist, fill = '#795695', color = '#795695') +
  geom_sf(data = occ_comb_sf, fill = "black", color = "black", size = 0.01) +
  guides(fill = guide_legend(title = "Sea Surface Temperature (C)")) 

```

