---
title: "Lab 3 Assignment - Scale"
output: html_notebook
---


## Challenge 1 (4 points)

**Build a raster with 100 rows and 100 columns. Fill the raster cells with values of a random variable drawn from a distribution of your choosing (Poisson, Normal, Uniform, etc.). Calculate the mean and variance of the values in that raster. Now increase the grain size of those cells by factors of 2, 5, and 10, combining cell values using a mean function. At each iteration, calculate the mean and variance of the values in the resulting raster. Generate 2 scatterplots that have grain size on the x-axis. Plot the mean raster value on the y-axis of the first, and variance on the y-axis of the second. What do you notice about how these values change as you "scale up" the grain size? Why do you think this pattern occurs?**

Place your code in the code chunk below so I can reproduce your analyses/figures.

```{r}
#requiring necessary packages
require(tidyverse)
require(sf)
require(AICcmodavg)
require(tigris)
require(FedData)
require(terra)
require(tidyterra)

#build raster (100x100)
simpRast = rast(ncol=100, nrow=100, xmin=1, xmax=100, ymin=1, ymax=100)

#fill raster (random variable drawn from distribution)
set.seed(2)
simpRast[] = rnorm(ncell(simpRast), mean=25, sd=3)

plot(simpRast)
text(simpRast)

#calculate mean and variance of values in raster
global(simpRast, mean) # Mean is 25.0331	
global(simpRast, var) #var is 8.994168	

#increase grain size by factor of 2, 5, and 10 using mean function
simpRastI2 <- aggregate(simpRast, fact=2, fun='mean')#mean value
simpRastI5 <- aggregate(simpRast, fact=5, fun='mean')#mean value
simpRastI10 <- aggregate(simpRast, fact=10, fun='mean')#mean value


#calculate mean and variance of values in resulting rasters from previous step
means <- c(mean(as.matrix(simpRast)), mean(as.matrix(simpRastI2)), mean(as.matrix(simpRastI5)), mean(as.matrix(simpRastI10)))
variances <- c(var(as.matrix(simpRast)), var(as.matrix(simpRastI2)), var(as.matrix(simpRastI5)), var(as.matrix(simpRastI10)))
grain <- c(1, 2, 5, 10)
df <- data.frame(grain, means, variances)
  
#scatterplots (x-axis = grain size), y-axis = mean (plot 1) & variance (plot 2)
plot(grain, means)
plot(grain, variances)

```

As the grain size increases, there is no change in the mean values, however there is a decrease in the variance. The static mean value makes sense as the function used to increase the grain size is the mean, and so as you increase the grain size using the means, it shouldn't change the mean of the dataset as a whole. The variance decrease with increasing grain size because taking the mean of a subset of the data is going to naturally decrease the variance, and as that subset of data gets larger and reflects a larger portion of the whole dataset (i.e. a grain that includes 10 units rather than 1 or 2), the values are going to get closer to the true mean and thus decrease the variance seen between each "subset". 

$\color{red}{\text{Awesome. +4}}$

## Challenge 2 (4 points)

**Identify a situation in which you might use a summary function other than the mean to calculate new cell values when you scale up the grain of a raster (e.g., median, mode, minimum, maximum, etc.). Repeat the effort from Challenge 1 using this alternate function. Again, create two scatterplots showing how the mean and variance values of the raster change as you scale up the cell size by factors of 2, 5, and 10. Do you see a similar pattern? Compare and contrast your findings with those from Challenge 1.**

*Hint: You should be able to recycle your code from Challenge 1 with only a couple of small tweaks to answer this question.*

Place your code in the code chunk below so I can reproduce your analyses/figures.

```{r}
#build raster (100x100)
simpRast = rast(ncol=100, nrow=100, xmin=1, xmax=100, ymin=1, ymax=100)

#fill raster (random variable drawn from distribution)
set.seed(2)
simpRast[] = rnorm(ncell(simpRast), mean=25, sd=3)

plot(simpRast)
# text(simpRast)

#calculate mean and variance of values in raster
global(simpRast, mean) # Mean is 25.0331	
global(simpRast, var) #var is 8.994168	

#increase grain size by factor of 2, 5, and 10 using mean function
simpRastI2 <- aggregate(simpRast, fact=2, fun='modal')#mean value
simpRastI5 <- aggregate(simpRast, fact=5, fun='modal')#mean value
simpRastI10 <- aggregate(simpRast, fact=10, fun='modal')#mean value


#calculate mean and variance of values in resulting rasters from previous step
means <- c(mean(as.matrix(simpRast)), mean(as.matrix(simpRastI2)), mean(as.matrix(simpRastI5)), mean(as.matrix(simpRastI10)))
variances <- c(var(as.matrix(simpRast)), var(as.matrix(simpRastI2)), var(as.matrix(simpRastI5)), var(as.matrix(simpRastI10)))
grain <- c(1, 2, 5, 10)
df <- data.frame(grain, means, variances)
  
#scatterplots (x-axis = grain size), y-axis = mean (plot 1) & variance (plot 2)
plot(grain, means)
plot(grain, variances)

```

Here we see a different trend in means, but the same trend with variance. As we take the "mode" for each subset, this isn't an accurate representation of the true mean of the subset, and so by taking the mode for each subset we move further away from the true mean of the dataset represented in our new, larger clusters. The variance the same trend as in question 1, because as we try to combine data into any type of summary statistics, we are naturally taking out variation seen and thus the variance of the overall dataset after summarizing it into different cluster sizes naturally decreases, and the fewer datapoints we have in the end, the less variance we are going to see. 

$\color{red}{\text{Ok, but why is the mean value decreasing? And when might you use the mode instead of the mean to summarize a raster? +3.5}}$

## Challenge 3 (2 points)

**Recall that before we calculated forest cover, we cropped our NLCD raster to minimize its size and the computing effort necessary from our poor little computers. How might that affect our ability to evaluate the scale at which five-lined skinks respond to forest cover? Why?**

In the example we cropped the NLCD raster to our study area variable after adding 10km in all directions from the extent of points sampled. This necessitates that at the points near the edges of this study area, if we try to look at a scale larger than 10 km they will all innately be 0 because we didn't include them in our nlcd raster of forest cover. 

$\color{red}{\text{Perfect. +2}}$

## Challenge 4 (4 points)

**In the lab, we measured forest cover at 1 km and 5 km. Extract forest cover proportions around each sample point for 100 m, 500 m, 1 km, 2 km, 3 km, 4 km, and 5 km scales. Examine the correlation between these 7 variables (remember the chart.Correlation() function). What patterns do you notice in correlation among these variables?**

*Hint: Recall the for loop we used to calculate this variable at two scales... could you make a small addition here to look at more scales?*

```{r}
# Including all the commands used in the lab exercise which are necessary to be able to extract forest cover proportions at different scales - i.e. removing demonstration sections and plotting sections, but otherwise only changing code to add the extra scales indicated

sites = st_read("/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week3/reptiledata.shp") %>% 
  filter(management!='Corn')
st_crs(sites) = "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
head(sites)


states = states() %>% 
  filter(NAME %in% c('Alabama', 'Florida', 'Georgia')) %>% 
  st_transform(crs(sites, proj=T))

presAbs = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week3/reptiles_flsk.csv')

sites = sites %>% 
  left_join(presAbs, by='site')

#Extract x and y coordinates of the bounding box
studyArea = st_bbox(sites) + c(-10000, -10000, 10000, 10000)
studyArea = st_as_sfc(studyArea)

nlcd = get_nlcd(studyArea,
                label='studyArea',
                year = 2016,
                dataset = 'landcover',
                landmass = 'L48'
)

forest = nlcd %>% 
  setValues(0)

forest[nlcd=='Deciduous Forest' | nlcd=='Evergreen Forest' | nlcd=='Mixed Forest'] = 1


# JV's function
bufferCover = function(shp, size, landcover){
  buffArea = (pi*size^2)/10000
  grainArea = (prod(res(landcover)))/10000
  
  buffi = st_buffer(shp[i,], dist=size)
  cropi = crop(landcover, buffi, mask=T)
  numCells = global(cropi, 'sum', na.rm=T)
  forestHa = numCells * grainArea
  propForest = forestHa / buffArea
  
  return(propForest)
}


#Here is where we are setting our different scales
#Extract forest cover proportions around each sample point for 100 m, 500 m, 1 km, 2 km, 3 km, 4 km, and 5 km scales

#This is where we are going to store the output values
for100m = as.vector(rep(NA, nrow(sites)))
for500m = as.vector(rep(NA, nrow(sites)))
for1km = as.vector(rep(NA, nrow(sites)))
for2km = as.vector(rep(NA, nrow(sites)))
for3km = as.vector(rep(NA, nrow(sites)))
for4km = as.vector(rep(NA, nrow(sites)))
for5km = as.vector(rep(NA, nrow(sites)))

for(i in 1:nrow(sites)){
  for100m[i] = bufferCover(sites, 100, forest)
  for500m[i] = bufferCover(sites, 500, forest)
  for1km[i] = bufferCover(sites, 1000, forest)
  for2km[i] = bufferCover(sites, 2000, forest)
  for3km[i] = bufferCover(sites, 3000, forest)
  for4km[i] = bufferCover(sites, 4000, forest)
  for5km[i] = bufferCover(sites, 5000, forest)
}

forestData = sites %>% 
  mutate(for100m = unlist(for100m),
         for500m = unlist(for500m),
         for1km = unlist(for1km),
         for2km = unlist(for2km),
         for3km = unlist(for3km),
         for4km = unlist(for4km),
         for5km = unlist(for5km))

forestData %>% 
  as.data.frame() %>% 
  select(coords_x1, for100m, for500m, for1km, for2km, for3km, for4km, for5km) %>% 
  PerformanceAnalytics::chart.Correlation(histogram=F)

```
As the scales get larger they become more correlated to the one lower, i.e. 4km x 5km is the highest correlation, whereas comparing 1km x 2km is still correlated, but not near as strong as 4x5. 

$\color{red}{\text{Specifically, they are positively correlated with one another. In addition, you'll notice that measurements further apart are less correlated with one another. +3}}$

## Challenge 5 (4 points)

**Fit 8 logistic regression models (a null model and one for each of the 7 forest scales). Compare these models using AICc. Which scale do you think represents the critical or characteristic scale at which forest cover affects skink presence? Is this scale clearly better than the others, or is there some ambiguity? What are some mechanisms by which forest cover could affect skink presence at this scale? What is your overall conclusion regarding how forest cover affects skink presence (i.e., take a look at the betas)?**

Place your R code in the chunk below.
```{r}

modelNull = glm(pres~1, family='binomial', data=forestData)
model100m = glm(pres~for100m, family='binomial', data=forestData)
model500m = glm(pres~for500m, family='binomial', data=forestData)
model1km = glm(pres~for1km, family='binomial', data=forestData)
model2km = glm(pres~for2km, family='binomial', data=forestData)
model3km = glm(pres~for3km, family='binomial', data=forestData)
model4km = glm(pres~for4km, family='binomial', data=forestData)
model5km = glm(pres~for5km, family='binomial', data=forestData)

aictab(list(modelNull, model100m, model500m, model1km, model2km, model3km, model4km, model5km), modnames=c('Null', '100 m', '500 m', '1 km', '2 km', '3 km', '4 km', '5 km'))

```
**Which scale do you think represents the critical or characteristic scale at which forest cover affects skink presence? Is this scale clearly better than the others, or is there some ambiguity? What are some mechanisms by which forest cover could affect skink presence at this scale? What is your overall conclusion regarding how forest cover affects skink presence (i.e., take a look at the betas)?**

Based on AICc values, the 2km scale is the best model (lowest AIC) to explain the data given the models used. The Delta AIC indicates that the models for 2-5km are all decent models, so there may be a bit of ambiguity (all have a delta AIC lower than 2), but overall the 2 km scale model is the best supported model of the models used. Forest cover could affect skink presence at a 2km scale as this could potentially be their average "home range size" - but I have no idea if that is true. Another possible explanation for 2km being the characteristic scale is that forest cover at this scale may impact/drive invertebrate abundance, which is the primary food source of five-lined skinks. Overall, while this 2km scale is the best model, I think all of the 3-5 km scales are largely correlated with one another and all are also supported as explaining the data of skink presence/absence, and could be interpreted together as all impacting skink presence. 

$\color{red}{\text{Specifically, skink presence increases with increasing forest cover across scales. +3.5}}$

## Challenge 6 (2 points)

**If you encounter ambiguity in identifying the characteristic scale of an effect, can you come up with a clever way to condense the information in the multi-scale variables into just one or two? When might it be ok to include two covariates in the same model (think multiple regression) that represent the same ecological feature measured at different scales (e.g., forest cover at 1 km AND forest cover at 5 km in the same model)? I can think of both a biological and a statistical answer to this question.**


From a statistical sense, I suppose if there is NOT a correlation between the two spatial scales being used, then statistically there wouldn't be an issue with including both. Biologically, I suppose that if these two spatial scales represent different life-history aspects, i.e. if the 1km scale encompassed their normal range, and the 5km scale encompassed a regular migration of the species to breeding sites or something similar, then it would make sense to include both separately in your model (again assuming that the two variables are not correlated, because even if there is biological reasoning for this, I think that would statistically not be kosher). 

$\color{red}{\text{I think you're saying that your clever way to condense information into two scales would be to assess the biological scales at which the species is likely to respond to forest cover, ensure they are not correlated, then choose those two scales for your model. +2}}$
