---
title: "Lab 5 assignment - Point pattern analysis"
output:
  html_document:
    df_print: paged
---

```{r}
rm(list=ls())

require(spatstat)
require(tidyverse)
require(terra)
```


# Challenge 1 (3 points)

Create a simulation window on a landscape stretching from 0 to 100 units in both the x and y directions. Now simulate the distribution of an organism on this landscape using a Poisson process with a lambda value of 0.01. Plot this organism's distribution. How many organisms are there on your landscape? What is the expected number of organisms that you should find on this landscape based on your simulation parameters? Why?

```{r}
set.seed(45)
Window = owin(xrange = c(0, 100), yrange = c(0, 100))

simPp = rpoispp(lambda = 0.01, nsim=1, win=Window)
plot(simPp)
simPp
```

There are 103 organisms on my landscape. The expected number is about 100 organisms because the intensity (or lambda) of points is set to 0.01, so for every square m you expect 0.01 organisms, and you can multiply that by the total area (here 100x100, so 10,000) to get the expected number of individuals (here 10,000 x 0.01 = 100).


# Challenge 2 (3 points)

Verify that there no linear or quadratic pattern in the distribution of your organisms using fitted Poisson process models. Show the outcome of this analysis and explain your conclusion.

```{r}
#Place your code here.
ppInt = ppm(simPp, ~1)
ppXy = ppm(simPp, ~ x + y)
ppXy2 = ppm(simPp, ~polynom (x, y, 2))


data.frame(model = c('int', 'xy', 'xy2'),
           AIC = c(AIC(ppInt), AIC(ppXy), AIC(ppXy2)))

```

Here I fitted the data to homogeneous (int), linear (xy), and quadratic(xy2). The best fitting model was the homogeneous, which indicates that there is no linear or quadratic pattern in the simulated data, rather it supports that the data is homogeneous.  


# Challenge 3 (14 points)

I have simulated the distributions of 2 critters and provided their x-y coordinates. One of them is a non-territorial predator. The second is a critter that is preyed upon by the predator and exhibits a fear response to the predator's presence. Use the tools you learned in the lab to explore the distributions of these two critters. Identify which one is the predator and which is the prey, and provide code and a thorough justification supporting your conclusion.


```{r}
#Use this code chunk to get you started.

predPreyWin = owin(xrange = c(0, 100), yrange = c(0, 100))

critter1 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species1.csv') %>% 
  mutate(org = 'critter1')
pppcritter1 = ppp(x=critter1$x, y=critter1$y, window = predPreyWin)

critter2 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species2.csv') %>% 
  mutate(org = 'critter2')
pppcritter2 = ppp(x=critter2$x, y=critter2$y, window = predPreyWin)

critters = rbind(critter1, critter2)
pppcritters = ppp(critters$x, critters$y, window = predPreyWin, marks = critters$org)


# Map distributions
plot(pppcritter1)
plot(pppcritter2)
plot(pppcritters)
#honestly these both look splotchy and aggregated, and neither look entirely random. My initial thought is that critter 1 is the prey, largely because of the large clump in the top left. 
```

From the plots it is difficult to discern which is randomly distributed and which is not. Both appear to be clumping a little bit, however my guess is that critter 1 is the prey, largely because of the large clump in the top left. This really does show, however, that you cannot tell random distributions just by looking (or at least apparently I cannot). When plotting both together (3rd plot) you can see a bit better that the triangles (critter 2) are more evenly spaces, whereas the circles seem to be more clumped together. Again, this is just by eye and not a good measure.

Now I am going to run some actual tests of randomness within each species. 


```{r}
#First test to look at randomness using quadrats

#use quadrats to look at randomness of critter 1
q = quadratcount(pppcritter1, nx=4, ny=4)
plot(pppcritter1)
plot(q, add=T)
quadrat.test(pppcritter1, nx=4, ny=4, method='Chisq')
#significant p-value in quadrat test indicates that this is significantly different from random. 

#use quadrats to look at randomness of critter 2
q = quadratcount(pppcritter2, nx=4, ny=4)
plot(pppcritter2)
plot(q, add=T)
quadrat.test(pppcritter2, nx=4, ny=4, method='Chisq')

#p-value not significant, so it is not significantly different from random

```

Here the quadrat test is showing that when breaking the landscape up into 16 squares (4x4), some squares appear to have more points than expected under random conditions in critter 1, but not in critter 2. From the R-console you can see that critter 1 is significantly different from random, whereas critter 2 is not. For some reason the script is acting weird and when I run this section as a chunk the last R console with output from the second quadrat test doesn't pop up, but you can click on it and manually run it - the code itself works fine.

So overall it is looking already like critter 1 is non-randomly distributed, and critter 2 is randomly distributed. 

```{r}

#Critter 1 Ripley's L
Ltrans = Lest(pppcritter1, correction='translate')
plot(Ltrans, .-r~r, main="Critter 1 Ripley's L")

# Critter 2 Ripley's L
Ltrans = Lest(pppcritter2, correction='translate')
plot(Ltrans, .-r~r, main="Critter 2 Ripley's L")

set.seed(55)

#Now testing for statistical difference under CSR
lCsr = envelope(pppcritter1, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(lCsr, .-r~r, shade=c('hi', 'lo'), legend=F, main="Critter 1 CSR Ripley's L")

lCsr = envelope(pppcritter2, Lest, nsim=99, rank=1, correction='translate', global=F)
plot(lCsr, .-r~r, shade=c('hi', 'lo'), legend=F, main="Critter 2 CSR Ripley's L")
```
From these true values of Ripleys L there is some deviation from expected seen with the dashed red line, however looking at the y-axis these deviation values are quite low. When plotting our observed pattern against simulated complete spatial randomness in the landscape (lCsr) we see that Critter 1 slightly falls outside of the confidence interval at lower range scales, however Critter 2 is largely within the confidence interval. This is indicating that for critter 1, when examining circular areas with smaller radii (at smaller scales) there are slightly more individuals of critter 1 than expected under random, but that trend goes away as you increase the radius of the circles measured. 

This particular line of evidence isn't selling me, but it doesn't entirely negate previous findings either. 

Given the evidence so far, I still think critter 1 is the non-random prey that is avoiding a randomly distributed predator that it fears, but lets look at how critter 1 is distributed relative to critter 2. 

I'm going to do one last set of tests to see how the animals are distributed relative to each other. 

```{r}
critter_name <- factor(pppcritters$marks)
#not sure why but I was getting an error that the point pattern must be multitype and this works. Probs could also have run factor before coverting to ppobject and that would probably work
marks(pppcritters) <- factor(pppcritters$marks) 

# Critter 1 to critter 1 (i.e. is critter 1 randomly distributed amongst itself)
lMulti = envelope(pppcritters, Lcross, nsim=99, rank=1, global=F, i="critter1", j="critter1", simulate = expression(rlabel(pppcritters)))
plot(lMulti, .-r~r, legend=F, main="Critter 1 - Critter 1 Lcross")
# Critter 2 to critter 2 (i.e. is critter 2 randomly distributed amongst itself)
lMulti = envelope(pppcritters, Lcross, nsim=99, rank=1, global=F, i="critter2", j="critter2", simulate = expression(rlabel(pppcritters)))
plot(lMulti, .-r~r, legend=F, main="Critter 2 - Critter 2 Lcross")

# Critter 1 to critter 2 (i.e. is critter 1 randomly distributed relative to critter 2)
lMulti = envelope(pppcritters, Lcross, nsim=99, rank=1, global=F, i="critter1", j="critter2", simulate = expression(rlabel(pppcritters)))
plot(lMulti, .-r~r, legend=F, main="Critter 1 - Critter 2 Lcross")

```

The results of these three plots indicates that 1. there are more critter 1s relative to critter 1 than would be expected under random, which means they are clumping together, 2. there are an expected number of critter 2s relative to critter 2 under random, which means critter 2 is  randomly distributed, and 3. when looking at critter 1 relative to critter 2, we see a significant dip below expected under random, indicating that critter 1 is further away from critter 2 than expected by random chance. This Lcross test provides the final support I need to conclude that critter 1 is indeed the prey that is actively avoiding a predator which is resulting in the critter 1 being slightly clustered together on the landscape, and critter 2 is a randomly distributed predator that those prey are avoiding. 
